---
title: "Some simple rank 0 examples"
author: "Matthew Stephens"
date: "September 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A single example

First we will simulate some rank 0 data:

```{r}
  library("flashr")
  set.seed(1)
  n=100
  p=1000
  Y = matrix(rnorm(n*p),nrow=n)
```


Run flash to fit a rank 1 model (`Kmax=1`). Here `nullcheck=FALSE` tells flash to keep
the final value of the variational optimization without checking
it against the rank 0 solution (since we are going to do that
by hand here.)
```{r}
  data = flash_set_data(Y)
  f1= flash(data,Kmax=1,nullcheck=FALSE)
  flash_get_objective(data,f1)
```

Now try deleting the factor - that gives us a rank 0 solution in this case. We can see here that deleting the factor actually increases
the objective function! That is the optimization did not
work very well (because deleting the factor is actually a special
case of the model). 
```{r}
f0 = flash_zero_out_factor(data,f1,1)
f0 = flash_update_precision(data,f0)
flash_get_objective(data,f0)
```

This is actually quite a common occurance (see below).
So we have implemented this check as the default in flash.
Here is what happens if you run `flash` with default (`nullcheck=TRUE`): it automatically deletes the factor.
```{r}
f1 = flash(data,Kmax=1,verbose=TRUE)
```


# Small simulation study: 

Let's see how often this happens:
```{r}
set.seed(1)
n=100
p=1000
F1minusF0 = rep(0,100)
for(i in 1:100){
  Y = matrix(rnorm(n*p),nrow=n)
  data = flash_set_data(Y)
  f1= flash(data,Kmax=1,nullcheck=FALSE)
  f0 = flash_zero_out_factor(data,f1,1)
  f0 = flash_update_precision(data,f0)
  F1minusF0[i] = flash_get_objective(data,f1) - flash_get_objective(data,f0)
}
hist(F1minusF0)
```



