---
title: "More advanced flash usage"
author: "Matthew Stephens"
date: "September 23, 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flashr advanced}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment = "#",collapse = TRUE,
  fig.align = "center",tidy = FALSE)
```

# Introduction

This vignette is designed to describe some more advanced use of flash.
Most users should start with the [flash_intro.html](flash_intro.html)
vignette.

The interface described may change as we develop `flashr`.

# Simulate Data

We start by simulating some data to provide an example:

```{r}
library(flashr)
set.seed(1)
n = 100
p = 500
k = 7
LL = matrix(rnorm(n*k),nrow=n)
FF = matrix(rnorm(p*k),nrow=p)
Y = LL %*% t(FF) + rnorm(n*p)
```

# Setting up data

For anything beyond the simplest usage, you should start your analysis
by setting up the data in what we will call a "flash data object".
This is essentially simply an n by p matrix, but with meta-data to
allow `flash` functions to deal properly with missing values
etc. (This framework also makes it easier for future extensions to
pass more information in as data; e.g. maybe variances also associated
with the matrix).

```{r}
data = flash_set_data(Y)
```

# Initializing a factor model

Second, we need to initialize a factor model - essentially a set of
factors and loadings. We call this a "flash fit object", although at
this point it is not actually fit to the data (this is the next step!)
There are several ways to do this, using functions of the form
`flash_add_xxx`. They are called this because they can add factors to
an existing flash fit object---by providing a value for the argument
`f_init`---so you can build up a flash fit object bit-by-bit. Here
we have no existing flash fit object, so we do not specify `f_init`,
and in this case these functions create a new fit object.

For illustration we show `flash_add_factors_from_data`. This
essentially runs `softImpute` (soft-thresholded singular value
decomposition) on the data to obtain an initial set of factors and
loadings. We have to choose a number of factors to initialize
with. (Factors can be dropped out during the fit so ideally this
should be a larger number than we think we actually need; but at the
same time more factors mean more computation, so you might start with
a moderate number, and then try increasing later if the data seem to
warrant it.)

```{r}
f = flash_add_factors_from_data(data,K=10)
```

# Refining/fitting a factor model

Once you have initialized a flash fit object, you can improve the fit
using `flash_backfit`. (If you want to see how the fit is progressing,
set `verbose=TRUE` here.)

```{r}
f = flash_backfit(data,f,verbose=FALSE)
```

# Extracting information

Having fit the model we might want to extract the fitted factors and
loadings.  You can do that using functions like `flash_get_ldf` to get
the standardized loadings, factors and weights. Also `flash_get_lf`
gets the product (LDF').  And `flash_get_objective` will return the
measure of goodness-of-fit achieved (the variational lower bound,
F). This can be helpful for comparing multiple fits to the same data.

```{r, fig.width = 5, fig.height = 5}
plot(flash_get_lf(f), LL %*% t(FF),
     main="compare fitted values with true LF'")
flash_get_ldf(f)$d
flash_get_objective(data,f)
```

# Greedy addition of factors

The way we did it above was to add all the factors at once and then optimize.
An alternative is to add one at a time, and optimize each in turn. This is accomplished
using `flash_add_greedy`. This will keep adding factors until they no longer improve
the fit, so you need to specify a maximum number to consider.
```{r}
f_greedy = flash_add_greedy(data,Kmax=10)
flash_get_ldf(f_greedy)$d
flash_get_objective(data,f_greedy)
```

After adding factors in this way you can use backfitting to try to
further improve the fit. In our experience, for different datasets,
sometimes this provides a better fit than initializing all at once (as
in `f` above), sometimes worse.

```{r}
f_greedy_bf = flash_backfit(data,f_greedy)
flash_get_ldf(f_greedy_bf)$d
flash_get_objective(data,f_greedy_bf)
flash_get_objective(data,f)
```

# Setting loadings and factors directly

In simulation experiments it might be useful to initialize the
loadings and factors to the "truth" to see how it affects the
convergence.

```{r}
f_true = flash_add_lf(data, LL=LL, FF=FF)
f_cheat = flash_backfit(data,f_true)
flash_get_ldf(f_cheat)$d
flash_get_objective(data,f_cheat)
```

Unexpectedly (to me) the objective achieved from this initialization
is lower than those from the greedy initialization. However, the
results are pretty similar in terms of mean squared error
(MSE). Notice that backfitting improved MSE vs greedy.

```{r}
mean((flash_get_lf(f_cheat) - LL %*% t(FF))^2)
mean((flash_get_lf(f_greedy_bf) - LL %*% t(FF))^2)
mean((flash_get_lf(f) - LL %*% t(FF))^2)
mean((flash_get_lf(f_greedy) - LL %*% t(FF))^2)
```

# Setting the function used to solve the EBNM problem

The flash algorithm involves iteratively applying a function that
solves the "Empirical Bayes normal means" (EBNM) problem. See Wang and
Stephens (2017) for more details on this.  The `flashr` package
provides two functions to do this: `ebnm_pn`, which uses a
point-normal prior from the `ebnm` package, and `ebnm_ash` which
provides an interface to functions in the `ashr` package.

In principle the `ebnm_ash` function fits a more flexible model, so
might be preferred. However, in the paper we generally found the two
methods give similar average performance, and `ebnm_pn` is (currently)
faster, and so this is the default.

```{r}
f_greedy_ash = flash_add_greedy(data,Kmax=10,ebnm_fn = ebnm_ash)
flash_get_objective(data,f_greedy_ash)
```

# Fixed factors or loadings

Sometimes you might want to include a "fixed" factor or loading in the
analysis.  For example, maybe you have a covariate you are interested
in the effect of.  Or you want to include a mean term in the rows or
columns. You can add such "fixed" factors using `flash_add_fixed_l` or
`flash_add_fixed_f`.  Fixed loadings will not be updated during a
subsequent fit, but their corresponding factors *will* be
updated. Similarly, fixed factors will not be updated, but their
loadings will be. (Missing values, NA, in a fixed loading or factor
are allowed and will be updated during a fit.)

For example, the following creates a flash fit object with a fixed
intercept loading. (So when fit it will estimate a corresponding
factor, which should be interpreted as a column-specific mean.) Then
it adds 10 data-based factors and loadings, and then it fits the
model. Note how the first loading (the mean) does not change during
the fit. (Notice also that the corresponding factor is 0---correct
since we did not add a non-zero mean to the simulation.)

```{r}
f = flash_add_fixed_l(data,LL = cbind(rep(1,n)))
f = flash_add_factors_from_data(data,K=10,f_init=f)
f = flash_backfit(data,f)
flash_get_k(f) #tells you how many factors f has
head(flash_get_ldf(f,1,drop_zero_factors = FALSE)$l)
# Note this is 0 as the factor has no weight
flash_get_ldf(f,1,drop_zero_factors = FALSE)$d
```
