---
title: "Flash introduction"
author: "Matthew Stephens"
date: "September 23, 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to flashr2

This is a brief introduction to the `flashr2` package, which is an
in-progress refactoring of the original `flashr` code. It is aimed
at people who might want to see some of the more "advanced" options.
We will want a simple single function interface for novice users.

All this is in very active development, and functionality/interface subject to change
without notice. You have been warned!

## Overview

First we will simulate some data for illustration (in this case the loadings
and factors are not at all sparse, so the benefits of flash over other methods might be limited.)
```{r}
library(flashr2)
set.seed(1)
n = 100
p = 500
k = 7
LL = matrix(rnorm(n*k),nrow=n)
FF = matrix(rnorm(p*k),nrow=p)
Y = LL %*% t(FF) + rnorm(n*p)
```


There are three steps to a flash analysis. 

First, set up the data in what we will call a "flash data object".
This is essentially simply an n by p matrix. (We do it this way so it will be easy 
for future extensions to pass more information in as data; e.g. maybe variances
also associated with that matrix).
```{r}
data = flash_set_data(Y)
```

Second, we need to initialize a factor model - essentially a set of factors
and loadings. We call this a "flash fit object", although at this point it is not 
actually fit to the data (this is the next step!) There are several ways to do this,
using functions of the form `flash_add_xxx`. They are called this because they
can add factors to an existing flash fit object -- by providing a value
for the argument `f_init` --- so you can build up a flash fit
object bit-by-bit. Here we have no existing flash fit object, so we do not
specify `f_init`, and in this case these functions create a new fit object.

For illustration we use what will probably be the most common first choice of method: 
`flash_add_factors_from_data`. This essentially runs
`softImpute` (soft-thresholded singular value decomposition) on the data to 
obtain an initial set of factors and loadings. We have to choose a number of factors
to initialize with. (Factors can be dropped out during the fit so ideally this
should be a larger number than we think we actually need; but at the same time
more factors mean more computation, so you might start with a moderate number, and then try increasing later if the data seem to warrant it.)
```{r}
f = flash_add_factors_from_data(data,K=10)
```

Third, and finally, we need to fit the flash model to the data. This is
done using `flash_backfit`. (If you want to see how the fit is progressing, set `verbose=TRUE` here)
```{r}
f = flash_backfit(data,f,verbose=FALSE)
```

Having fit the model we might want to extract the fitted factors and loadings.
You can do that using functions like `flash_get_l` to get the loadings and
`flash_get_f` to get the factors. Also `flash_get_lf` gets the product (LF').
And `flash_get_pve` gives a rough summary of the proportion of variance in Y 
explained by each factor-loading combination. And `flash_get_F` will return the 
measure of goodness-of-fit achieved (the variational lower bound, F). This can
be helpful for comparing multiple fits to the same data.
```{r}
plot(flash_get_lf(f), LL %*% t(FF), main="compare fitted values with true LF'")
flash_get_pve(f)
flash_get_F(data,f)
```

# Greedy addition of factors

The way we did it above was to add all the factors at once and then optimize.
An alternative is to add one at a time, and optimize each in turn. This is accomplished
using `flash_add_greedy`. This will keep adding factors until they no longer improve
the fit, so you need to specify a maximum number to consider.
```{r}
f_greedy = flash_add_greedy(data,Kmax=10)
flash_get_pve(f_greedy)
flash_get_F(data,f_greedy)
```

After adding factors in this way it is a good idea to use backfitting to try
to further improve the fit. In this case we get a slightly better fit
than the original (although it maybe took longer?)
```{r}
f_greedy_bf = flash_backfit(data,f_greedy)
flash_get_pve(f_greedy_bf)
flash_get_F(data,f_greedy_bf)
```

# Setting loadings and factors directly

In simulation experiments it might be useful to initialize
the loadings and factors to the "truth" to see how it affects the convergence.
```{r}
f_true = flash_add_lf(data, LL=LL, FF=FF)
f_cheat = flash_backfit(data,f_true)
flash_get_pve(f_cheat)
flash_get_F(data,f_cheat)
```

Unexpectedly (to me) the objective achieved from this initialization
is lower than those from the other initializations. However, the
results are pretty similar in terms of mean squared error:
```{r}
mean((flash_get_lf(f_cheat) - LL %*% t(FF))^2)
mean((flash_get_lf(f_greedy_bf) - LL %*% t(FF))^2)
mean((flash_get_lf(f) - LL %*% t(FF))^2)
mean((flash_get_lf(f_greedy) - LL %*% t(FF))^2)
```



# Fixed factors or loadings

Sometimes you might want to include a "fixed" factor or loading in the analysis.
For example, maybe you have a covariate you are interested in the effect of.
Or you want to include a mean term in the rows or columns. You can
add such "fixed" factors using `flash_add_fixed_l` or `flash_add_fixed_f`.
Fixed loadings will not be updated during a subsequent fit, but their corresponding
factors *will* be updated. Similarly, fixed factors will not be updated, but
their loadings will be. (Missing values, NA, in a fixed loading or factor are allowed
and will be updated during a fit.)

For example, the following creates a flash fit object with a fixed intercept loading.
(so when fit it will estimate a corresponding factor, which should be interpreted as a column-specific mean). Then it adds 10 data-based factors and loadings, and then it fits the model. Note how the first loading (the mean) does not change during the fit.
(Notice also that the corresponding factor is 0 - correct since we did not add a non-zero mean to the simulation.) 
```{r}
f = flash_add_fixed_l(data,LL = cbind(rep(1,n)))
f = flash_add_factors_from_data(data,K=10,f_init=f)
f = flash_backfit(data,f)
flash_get_l(f,1)
flash_get_f(f,1)
flash_get_pve(f)
```


